---
title: "1. Evaluating forecasts: RAFE, C-RAFE and the Sharpe gap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Evaluating forecasts: RAFE, C-RAFE and the Sharpe gap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
options(digits = 4)
```

```{r}
library(rafe)
```

This vignette works through the evaluation example of Salcher, Stöckl & Hanke
(2026, *Journal of Forecasting*) on the 12 Fama-French industry portfolios.

The question the paper asks is simple. A forecaster hands you a mean vector
$\hat\mu$ and a covariance $\hat\Sigma$. You feed them into a mean-variance
optimiser. **How much Sharpe ratio do you lose because the forecasts were
wrong?** The conventional answer — look at RMSE — turns out to be the wrong
instrument, because RMSE does not know which errors the optimiser cares about.

## The setup

We roll a 60-month estimation window and a 60-month evaluation window across
the sample, stepping a year at a time. On each window the estimated moments
play the role of the *forecast* $(\hat\mu, \hat\Sigma)$ and the realised
moments of the following window play the role of the *truth* $(\mu, \Sigma)$.

```{r}
data(ff12)
R     <- as.matrix(ff12[, -1])
dates <- ff12$date
c(months = nrow(R), assets = ncol(R))
```

```{r}
T_train <- 60L
T_eval  <- 60L
step    <- 12L
starts  <- seq(1L, nrow(R) - T_train - T_eval + 1L, by = step)

# Sharpe ratio of the plug-in tangency portfolio, evaluated under the truth.
plugin_sharpe <- function(mu_hat, S_hat, mu, Sigma) {
  iSh <- solve(S_hat)
  w   <- drop(iSh %*% mu_hat) / drop(t(mu_hat) %*% iSh %*% mu_hat)
  drop(t(mu) %*% w) / sqrt(drop(t(w) %*% Sigma %*% w))
}

panel <- do.call(rbind, lapply(starts, function(s) {
  tr <- R[s:(s + T_train - 1L), ]
  ev <- R[(s + T_train):(s + T_train + T_eval - 1L), ]

  mu_hat <- colMeans(tr); S_hat <- stats::cov(tr)
  mu     <- colMeans(ev); Sigma <- stats::cov(ev)

  bound   <- compute_trafe(mu_hat, mu, S_hat, Sigma)
  SR_star <- attr(bound, "SR_star")

  data.frame(
    date    = dates[s + T_train],
    rafe    = attr(bound, "rafe"),
    crafe   = attr(bound, "crafe"),
    trafe   = as.numeric(bound),
    c_used  = attr(bound, "c"),
    SR_star = SR_star,
    SR_hat  = plugin_sharpe(mu_hat, S_hat, mu, Sigma),
    rmse    = sqrt(mean((mu_hat - mu)^2))
  )
}))
panel$delta <- panel$SR_star - panel$SR_hat

nrow(panel)
head(panel[, c("date", "rafe", "crafe", "trafe", "delta", "rmse")])
```

## Does the bound hold?

The theorem says the realised Sharpe gap $\Delta = SR^{*} - SR(\hat w)$ never
exceeds $c \cdot \mathrm{RAFE} + SR^{*} \cdot \mathrm{C\text{-}RAFE}$.

```{r}
c(violations = sum(panel$delta > panel$trafe + 1e-10),
  windows    = nrow(panel))

summary(panel$trafe / panel$delta)
```

No violations. Note also how *loose* the bound is in practice — the median
ratio of bound to realised gap is well above one. The bound's value is not that
it is tight; it is that its two components tell you *where* the loss comes
from, and that they are the things you can act on.

`compute_trafe()` applies the theorem's data-dependent constant automatically
($c = 1$ when $\mathrm{RAFE} \le SR^{*}$, else $c = 2$):

```{r}
table(c = panel$c_used)
```

## Why RMSE is the wrong instrument

Here is the paper's central empirical point. Both RMSE and RAFE measure the
error in $\hat\mu$. Only one of them tracks the economic damage.

```{r}
cor_tbl <- data.frame(
  metric      = c("RMSE", "RAFE", "C-RAFE", "T-RAFE"),
  correlation = c(cor(panel$rmse,  panel$delta),
                  cor(panel$rafe,  panel$delta),
                  cor(panel$crafe, panel$delta),
                  cor(panel$trafe, panel$delta)),
  rank_corr   = c(cor(panel$rmse,  panel$delta, method = "spearman"),
                  cor(panel$rafe,  panel$delta, method = "spearman"),
                  cor(panel$crafe, panel$delta, method = "spearman"),
                  cor(panel$trafe, panel$delta, method = "spearman"))
)
cor_tbl
```

```{r}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1))
plot(panel$rmse, panel$delta, pch = 19, col = "grey35",
     xlab = "RMSE of the mean forecast", ylab = expression(Delta ~ "(Sharpe gap)"),
     main = "RMSE vs economic loss")
abline(lm(delta ~ rmse, panel), col = "firebrick", lwd = 2)
plot(panel$rafe, panel$delta, pch = 19, col = "grey35",
     xlab = "RAFE of the mean forecast", ylab = expression(Delta ~ "(Sharpe gap)"),
     main = "RAFE vs economic loss")
abline(lm(delta ~ rafe, panel), col = "firebrick", lwd = 2)
par(op)
```

Same forecast errors, same evaluation windows. Re-weighting the error by
$\Sigma^{-1}$ — charging the forecaster for mistakes in directions the
optimiser actually leans on — is what turns a forecast-accuracy number into a
decision-relevant one. On this sample RAFE correlates with the realised Sharpe
gap at roughly $0.69$ against RMSE's $0.19$.

### A result worth being explicit about

The same table shows C-RAFE and T-RAFE correlating *negatively* with $\Delta$.
That is not a contradiction, and it is worth understanding before anyone uses
T-RAFE as a scoring rule.

T-RAFE is an **upper bound**, not an estimate of the gap. C-RAFE is an operator
norm, so it is governed by the single worst-distorted eigendirection of
$\hat\Sigma$ relative to $\Sigma$. In windows where the sample covariance is
close to singular, C-RAFE explodes — but the plug-in portfolio does not
necessarily lose a correspondingly large amount of realised Sharpe, because it
may simply not load on the offending direction. The bound stays valid (it never
binds), while the scalar it produces stops ranking windows by economic damage.

The practical reading: use the **decomposition**, not the scalar. RAFE is the
quantity that tracks the mean channel's damage and that a mean corrector can
act on; C-RAFE is the quantity a covariance cleaner can act on. Adding them
gives you a certificate, not a leaderboard.

## The two channels through time

`compute_trafe()` returns the decomposition, so you can see whether a bad
window was a *mean* problem or a *covariance* problem.

```{r}
op <- par(mar = c(4.2, 4.2, 2.4, 1))
matplot(panel$date,
        cbind(panel$delta, panel$rafe, panel$SR_star * panel$crafe),
        type = "l", lty = 1, lwd = 2, col = c("black", "steelblue", "darkorange"),
        xaxt = "n", xlab = "", ylab = "",
        main = "Realised gap and its two bounding channels")
axis(1, at = pretty(panel$date), labels = format(pretty(panel$date), "%Y"))
legend("topleft", bty = "n", lwd = 2,
       col = c("black", "steelblue", "darkorange"),
       legend = c(expression(Delta), "RAFE (mean channel)",
                  expression(SR^"*" %*% "C-RAFE (covariance channel)")))
par(op)
```

On this universe the covariance channel is the larger of the two in about 90%
of windows, and roughly four times the mean channel on average:

```{r}
c(mean_channel       = mean(panel$rafe),
  covariance_channel = mean(panel$SR_star * panel$crafe),
  share_cov_larger   = mean(panel$SR_star * panel$crafe > panel$rafe))
```

That is the empirical motivation for pairing a mean corrector with a covariance
cleaner rather than fixing $\hat\mu$ alone — the subject of
`vignette("rafe-post-processing", package = "rafe")`.

## Reference

Salcher, T., Stöckl, S., & Hanke, M. (2026). Lost in Translation?
Risk-Adjusting RMSE for Economic Forecast Performance. *Journal of
Forecasting*. \doi{10.1002/for.70134}
