---
title: "2. Post-processing forecasts: correcting the moments you were handed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Post-processing forecasts: correcting the moments you were handed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
options(digits = 4)
```

```{r}
library(rafe)
data(ff12)
R <- as.matrix(ff12[, -1])
```

`vignette("rafe-evaluation", package = "rafe")` showed that the Sharpe gap of a
plug-in portfolio is bounded by two channels: a mean error measured in the risk
metric (RAFE) and a covariance distortion measured in the operator norm
(C-RAFE). This vignette does something about them.

The premise is that you have already been handed $(\hat\mu, \hat\Sigma)$ — from
a sample estimate, a vendor model, a machine-learning pipeline — and you cannot
retrain it. What you *can* do is correct the moments before they reach the
optimiser. That is the subject of the moment-correction working paper cited
below.

## The two atomic correctors

`mu_rafe_stein()` shrinks the mean toward a target with intensity $\kappa$,
computing that intensity in the $\hat\Sigma^{-1/2}$-whitened space. Pass
`T_obs` for the positive-part James-Stein intensity implied by the sample size:

```{r}
window    <- R[1:60, ]
mu_hat    <- colMeans(window)
Sigma_hat <- stats::cov(window)

mu_js <- mu_rafe_stein(mu_hat, Sigma_hat, T_obs = 60)
attr(mu_js, "kappa")

round(rbind(raw = mu_hat, corrected = as.numeric(mu_js)), 4)[, 1:6]
```

The corrector pulls the cross-section toward its own grand mean while leaving
that grand mean untouched:

```{r}
c(raw = mean(mu_hat), corrected = mean(mu_js),
  raw_spread = sd(mu_hat), corrected_spread = sd(mu_js))
```

`sigma_crafe_floor()` lifts the small eigenvalues of $\hat\Sigma$ to a floor set
relative to the mean eigenvalue. Those small eigenvalues are exactly the
directions a mean-variance optimiser leverages hardest, and exactly where the
sample covariance is least trustworthy:

```{r}
lam_raw <- eigen(Sigma_hat, symmetric = TRUE)$values
lam_cln <- eigen(sigma_crafe_floor(Sigma_hat, 0.3), symmetric = TRUE)$values
round(rbind(raw = lam_raw, floored = lam_cln), 5)
c(condition_raw = max(lam_raw) / min(lam_raw),
  condition_floored = max(lam_cln) / min(lam_cln))
```

## Choosing $\kappa$ and $\tau$ from data

Neither intensity should be picked by hand. Both tuners split the training
window into an inner-training and an inner-validation block (40 + 20 months by
default), fit on the first and score on the second.

They differ in *what* they score. `sep_tune()` tunes each parameter against its
own channel — $\kappa$ on inner-validation RAFE, $\tau$ on inner-validation
C-RAFE. `joint_trafe_tune()` searches the full $(\kappa, \tau)$ grid against
inner-validation T-RAFE.

```{r}
kg <- seq(0, 1, by = 0.05)
tg <- seq(0, 0.5, by = 0.05)

fit_sep   <- sep_tune(window, kappa_grid = kg, tau_grid = tg)
fit_joint <- joint_trafe_tune(window, kappa_grid = kg, tau_grid = tg)

rbind(sep   = c(kappa = fit_sep$kappa,   tau = fit_sep$tau_rel),
      joint = c(kappa = fit_joint$kappa, tau = fit_joint$tau_rel))
```

`sep_tune()` returns the two loss profiles, so you can see whether the optimum
is interior or sitting on a grid edge:

```{r}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 1))
plot(kg, fit_sep$losses_kappa, type = "l", lwd = 2, col = "steelblue",
     xlab = expression(kappa), ylab = "inner-validation RAFE",
     main = "Mean channel")
abline(v = fit_sep$kappa, lty = 2)
plot(tg, fit_sep$losses_tau, type = "l", lwd = 2, col = "darkorange",
     xlab = expression(tau), ylab = "inner-validation C-RAFE",
     main = "Covariance channel")
abline(v = fit_sep$tau_rel, lty = 2)
par(op)
```

## Does correcting the moments shrink the bound?

This is the question the package can answer directly, and it is worth asking
before any portfolio is formed. We roll a 60-month training window, tune on it,
and measure both channels against the realised moments of the following 60
months.

```{r}
T_train <- 60L
T_eval  <- 60L
starts  <- seq(1L, nrow(R) - T_train - T_eval + 1L, by = 12L)

channels <- do.call(rbind, lapply(starts, function(s) {
  tr <- R[s:(s + T_train - 1L), ]
  ev <- R[(s + T_train):(s + T_train + T_eval - 1L), ]
  mu <- colMeans(ev); Sigma <- stats::cov(ev)
  fit <- sep_tune(tr, kappa_grid = kg, tau_grid = tg)
  data.frame(
    rafe_raw  = compute_rafe(colMeans(tr), mu, Sigma = Sigma),
    rafe_cor  = compute_rafe(fit$mu_tilde, mu, Sigma = Sigma),
    crafe_raw = compute_crafe(Sigma, stats::cov(tr)),
    crafe_cor = compute_crafe(Sigma, fit$Sigma_tilde),
    date      = ff12$date[s + T_train]
  )
}))

round(rbind(
  RAFE   = c(raw = median(channels$rafe_raw),  corrected = median(channels$rafe_cor),
             share_improved = mean(channels$rafe_cor  < channels$rafe_raw)),
  `C-RAFE` = c(raw = median(channels$crafe_raw), corrected = median(channels$crafe_cor),
             share_improved = mean(channels$crafe_cor < channels$crafe_raw))
), 3)
```

Both channels fall. The covariance channel improves in every single window and
its median drops by roughly a factor of three; the mean channel improves in
about four windows out of five.

Not on average, but window by window:

```{r, fig.height = 5.4}
op <- par(mfrow = c(2, 1), mar = c(3.2, 4.4, 2.2, 1))
matplot(channels$date, cbind(channels$crafe_raw, channels$crafe_cor),
        type = "l", lty = 1, lwd = 2, col = c("#9AA5B1", "#1f4e79"),
        xlab = "", ylab = "C-RAFE", main = "Covariance channel")
legend("topleft", c("uncorrected", "after sigma_crafe_floor()"), bty = "n",
       lwd = 2, col = c("#9AA5B1", "#1f4e79"))
matplot(channels$date, cbind(channels$rafe_raw, channels$rafe_cor),
        type = "l", lty = 1, lwd = 2, col = c("#9AA5B1", "#E8A33D"),
        xlab = "", ylab = "RAFE", main = "Mean channel")
legend("topleft", c("uncorrected", "after mu_rafe_stein()"), bty = "n",
       lwd = 2, col = c("#9AA5B1", "#E8A33D"))
par(op)
```

The covariance line never crosses: the floor helps in all `r nrow(channels)`
windows, and by the largest margin exactly when the uncorrected distortion
spikes. The mean channel is noisier, which is what you would expect of a
shrinkage estimator — it trades a small loss in calm periods for a large gain
in turbulent ones.

## Taking it to portfolios

```{r}
T_hold <- 12L
tan_w  <- function(mu, S) { w <- drop(solve(S) %*% mu); w / sum(w) }

reb <- seq(1L, nrow(R) - T_train - T_hold + 1L, by = T_hold)
bt <- lapply(reb, function(s) {
  tr <- R[s:(s + T_train - 1L), ]
  ho <- R[(s + T_train):(s + T_train + T_hold - 1L), ]
  sp <- sep_tune(tr, kappa_grid = kg, tau_grid = tg)
  jt <- joint_trafe_tune(tr, kappa_grid = kg, tau_grid = tg)
  ws <- list(
    plugin = tan_w(colMeans(tr), stats::cov(tr)),
    sep    = tan_w(sp$mu_tilde, sp$Sigma_tilde),
    joint  = tan_w(jt$mu_tilde, jt$Sigma_tilde),
    `1/N`  = rep(1 / ncol(R), ncol(R))
  )
  list(r = vapply(ws, function(w) as.numeric(ho %*% w), numeric(T_hold)),
       pars = c(k_sep = sp$kappa, t_sep = sp$tau_rel,
                k_jnt = jt$kappa, t_jnt = jt$tau_rel))
})

rets <- do.call(rbind, lapply(bt, `[[`, "r"))
ann  <- function(x) c(mean = mean(x) * 12, sd = stats::sd(x) * sqrt(12),
                      sharpe = mean(x) / stats::sd(x) * sqrt(12))
round(t(apply(rets, 2, ann)), 3)
```

The uncorrected plug-in tangency portfolio is the disaster the literature leads
you to expect: an annualised volatility above 200% and a Sharpe ratio near zero.

Correcting the moments cuts that volatility by about **85%**, from roughly 2.28
to 0.36 annualised. The Sharpe ratio improves too, but far more modestly --
from about 0.09 to 0.11 -- because the correction shrinks the mean as well as
the risk. Most of what the eigenvalue floor buys you here is the removal of
enormous leveraged positions along badly-estimated directions, not extra return
per unit of risk.

Plotted through time, the difference is not where you might expect it:

```{r}
cum <- apply(rets, 2, cumsum)
dates_bt <- ff12$date[unlist(lapply(reb, function(s)
  (s + T_train):(s + T_train + T_hold - 1L)))]

matplot(dates_bt, cum, type = "l", lwd = 2, lty = c(1, 2, 1, 1),
        col = c("#B4413C", "#1f4e79", "#4FA3A5", "#E8A33D"),
        xlab = "", ylab = "cumulative monthly excess return",
        main = "Corrected vs uncorrected plug-in portfolios")
abline(h = 0, col = "grey60")
legend("topleft", colnames(rets), bty = "n", lwd = 2, lty = c(1, 2, 1, 1),
       col = c("#B4413C", "#1f4e79", "#4FA3A5", "#E8A33D"))
```

`sep` is drawn dashed because it lies exactly on top of `joint`.

Read that chart carefully, because the naive reading is wrong. The uncorrected
plug-in portfolio **ends highest**. Almost all of it comes from a single
episode in 1981-82, visible as the near-vertical excursion: the strategy
first loses several times its capital and then makes it back.

```{r}
rbind(`worst monthly return` = round(apply(rets, 2, min), 3),
      `months below -100%`   = colSums(rets < -1),
      `cumulative sum`       = round(cum[nrow(cum), ], 2))
```

A monthly return of $-8.3$ is not a drawdown, it is ruin several times over: an
investor holding that portfolio is wiped out and never reaches the recovery.
The cumulative-sum convention above quietly assumes the position is refinanced
at constant notional every month, which is what lets the red line come back at
all. On a compounded basis the plug-in portfolio's wealth goes negative and the
path simply ends.

That is what the correction buys. It does not add return; it removes the
leverage that makes the mean-variance solution uninvestable. The worst month
falls from $-833\%$ to $-116\%$, and the annualised volatility from 2.28 to
0.36.

**And equal weighting still beats both**, at about 0.46. That is not a bug in
the vignette, and it is worth stating plainly: on a 12-asset industry universe
with unconstrained weights, `1/N` is a famously hard benchmark. The working
paper's claims rest on a different and much more careful design -- larger
cross-sections, non-overlapping inference, and comparison against Ledoit-Wolf
and nonlinear shrinkage rather than against a raw plug-in. Nothing here
reproduces them; see the paper.

## The separability claim

The two tuners search different objectives, so there is no algebraic reason for
them to agree. On this sample they do, in every rebalance:

```{r}
pars <- do.call(rbind, lapply(bt, `[[`, "pars"))
c(kappa_agree = mean(pars[, "k_sep"] == pars[, "k_jnt"]),
  tau_agree   = mean(pars[, "t_sep"] == pars[, "t_jnt"]))

round(rbind(kappa = summary(pars[, "k_sep"]), tau = summary(pars[, "t_sep"])), 3)
```

That is the practical content of the separability claim: the bound's
two channels can be tuned independently without loss, which turns a
two-dimensional grid search into two one-dimensional ones. If you are tuning on
a large cross-section, prefer `sep_tune()` — it is the cheaper of the two and,
here, selects identically.

One caveat on the grid. Relative floors are admitted on `[0, 0.5]`, which is
the package default. On this universe the selected $\tau$ sits on the upper
edge of that range in well over half the rebalances:

```{r}
c(tau_median = median(pars[, "t_sep"]),
  share_at_upper_edge = mean(pars[, "t_sep"] == max(tg)))
```

A boundary-binding parameter means the data wants more flooring than the grid
allows, so the reported optimum is a constrained one. Widening the grid here
does improve realised Sharpe — but it leaves the range the paper works in, so
the package will not do it silently.

## References

Salcher, L., Stöckl, S., & Hanke, M. (2026). Lost in Translation?
Risk-Adjusting RMSE for Economic Forecast Performance. *Journal of Forecasting*.
\doi{10.1002/for.70134}

Stöckl, S., Salcher, L., & Hanke, M. Post-Optimal Moment Correction for
Mean-Variance Portfolios. Working paper.
