---
title: "1. Evaluating forecasts: reproducing the Part 1 experiment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Evaluating forecasts: reproducing the Part 1 experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
options(digits = 4)
```

```{r}
library(rafe)
data(ff49)
```

This vignette follows the mean-variance experiment of Salcher, Stöckl & Hanke
(2026, *Journal of Forecasting*), whose replication code lives at
[github.com/sstoeckl/Lost_in_Translation_Replication](https://github.com/sstoeckl/Lost_in_Translation_Replication).

The design solves a problem that makes forecast evaluation hard: real
forecasters do not come with a dial marked "accuracy". So the paper builds one.

## Forecasts of known quality

Take the rolling 60-month mean as a baseline forecast, then blend in a sliver of
the *realised* return:

$$\hat\mu_t(\lambda) \;=\; (1 - \lambda)\,\bar r_{t-60:t-1} \;+\; \lambda\, r_t,
\qquad \hat\Sigma_t(\lambda) \;=\; (1-\lambda)^2\,\hat\Sigma_t$$

At $\lambda = 0$ there is no look-ahead. As $\lambda$ rises the forecast gets
better in a controlled, measurable way. The covariance is scaled by
$(1-\lambda)^2$ to keep the contamination internally consistent — and that
choice has a useful side effect, which we check below.

```{r}
R  <- as.matrix(ff49[, -1])
nm <- colnames(R)
h_wind     <- 60L
lambda_vec <- seq(0, 0.1, by = 0.01)   # the paper's coarse grid
no_samples <- c(10, 20, 30)
n_draws    <- 10L                      # the paper uses 50; trimmed for runtime

oos     <- (h_wind + 1L):nrow(R)
roll_mu <- t(vapply(oos, function(t) colMeans(R[(t - h_wind):(t - 1L), ]),
                    numeric(ncol(R))))
roll_S  <- lapply(oos, function(t) stats::cov(R[(t - h_wind):(t - 1L), ]))
c(industries = ncol(R), months = nrow(R), oos_months = length(oos))
```

## C-RAFE has a closed form here

Because $\hat\Sigma = (1-\lambda)^2\Sigma$ is a scalar multiple of the truth,
the operator norm collapses analytically:

$$\mathrm{C\text{-}RAFE}(\lambda) \;=\; \bigl\lVert \Sigma^{1/2}\,(1-\lambda)^{-2}\Sigma^{-1}\,\Sigma^{1/2} - I \bigr\rVert_2
\;=\; \left|\tfrac{1}{(1-\lambda)^2} - 1\right|$$

which is exactly `compute_crafe_lambda()` in the replication code. That makes a
sharp check on the package's numerical implementation:

```{r}
Sigma_chk <- stats::cov(R[1:120, 1:8])
chk <- vapply(lambda_vec, function(l)
  compute_crafe(Sigma_chk, (1 - l)^2 * Sigma_chk), numeric(1))
round(rbind(computed = chk,
            closed_form = abs(1 / (1 - lambda_vec)^2 - 1)), 8)
```

Identical. Note the argument order: `compute_crafe()` takes the **realised**
covariance first, matching the replication code so paper scripts run unchanged.

## The experiment

For each draw of $N$ industries we form the plug-in tangency portfolio at every
out-of-sample month, exactly as the paper does, with weights budget-normalised
to sum to one.

One algebraic consequence of the $(1-\lambda)^2$ scaling is worth noticing:
since $\hat\Sigma^{-1} = (1-\lambda)^{-2}\Sigma^{-1}$, the scalar cancels in
$w = \hat\Sigma^{-1}\hat\mu / \mathbf{1}^\top\hat\Sigma^{-1}\hat\mu$. The
covariance contamination therefore does not touch the weights at all — it acts
purely through C-RAFE, while the realised portfolio responds only to the mean
channel.

```{r}
set.seed(50)
draws <- lapply(no_samples, function(k)
  lapply(seq_len(n_draws), function(d) match(sample(nm, k), nm)))

rows <- list(); z <- 0L
for (si in seq_along(no_samples)) {
  for (d in seq_len(n_draws)) {
    idx  <- draws[[si]][[d]]
    Rd   <- R[oos, idx, drop = FALSE]
    rm_d <- roll_mu[, idx, drop = FALSE]
    Sinv <- lapply(roll_S, function(S) solve(S[idx, idx, drop = FALSE]))

    Sigma_d <- stats::cov(Rd)
    mu_d    <- colMeans(Rd)
    w_star  <- solve(Sigma_d) %*% mu_d
    SR_star <- (sum(w_star * mu_d) / sum(w_star)) /
      sqrt(drop(t(w_star) %*% Sigma_d %*% w_star) / sum(w_star)^2)

    for (l in lambda_vec) {
      FC <- rm_d * (1 - l) + Rd * l
      pr <- vapply(seq_along(oos), function(i) {
        wr <- drop(Sinv[[i]] %*% FC[i, ])
        sum((wr / sum(wr)) * Rd[i, ])
      }, numeric(1))
      z <- z + 1L
      rows[[z]] <- data.frame(
        n_ind   = no_samples[si], draw = d, lambda = l,
        sharpe  = mean(pr) / stats::sd(pr) * sqrt(12),
        MSFE    = mean((Rd - FC)^2),
        RAFE    = compute_rafe(colMeans(FC), mu_d, Sigma_d),
        C_RAFE  = abs(1 / (1 - l)^2 - 1),
        SR_star = SR_star
      )
    }
  }
}
acc <- do.call(rbind, rows)
acc$T_RAFE <- acc$RAFE + abs(acc$SR_star) * acc$C_RAFE

agg <- aggregate(cbind(sharpe, MSFE, RAFE, C_RAFE, T_RAFE) ~ n_ind + lambda,
                 acc, mean)
round(agg[agg$n_ind == 30, ], 5)
```

## What the metrics say

Read that table by column. As $\lambda$ rises:

- **MSFE falls monotonically.** By the conventional standard, the forecast is
  getting steadily better.
- **RAFE falls too** — the mean channel genuinely does improve.
- **C-RAFE rises from zero**, because the covariance is being corrupted at the
  same time.
- **T-RAFE rises.** The sum of the two channels says the *decision* is getting
  worse even as the forecast error gets smaller.

```{r}
s <- agg[agg$n_ind == 30, ]
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.4, 1))
plot(s$lambda, s$MSFE, type = "b", pch = 19, col = "firebrick", lwd = 2,
     xlab = expression(lambda), ylab = "MSFE", main = "What RMSE sees")
plot(s$lambda, s$T_RAFE, type = "b", pch = 19, col = "steelblue", lwd = 2,
     xlab = expression(lambda), ylab = "T-RAFE", main = "What the bound sees")
par(op)
```

Two metrics, same forecasts, opposite verdicts. That is the paper's title: the
forecast improvement is real, but it gets *lost in translation* on the way to
the portfolio, because the accuracy gain and the covariance damage arrive
together and only one of them shows up in RMSE.

```{r}
by_n <- do.call(rbind, lapply(no_samples, function(k) {
  s <- agg[agg$n_ind == k, ]
  data.frame(n_ind = k,
             cor_MSFE   = cor(s$MSFE,   s$sharpe),
             cor_RAFE   = cor(s$RAFE,   s$sharpe),
             cor_T_RAFE = cor(s$T_RAFE, s$sharpe))
}))
round(by_n, 3)
```

## An honest caveat on the portfolio column

The realised Sharpe column is noisy and does not move monotonically in
$\lambda$, and the correlations above are correspondingly weak and unstable
across $N$. This is a property of the portfolio construction, not of the
metrics.

Budget normalisation divides by $\mathbf{1}^\top\hat\Sigma^{-1}\hat\mu$, which
for an unconstrained tangency portfolio on 10-30 industries passes close to zero
at some dates. When it does, leverage explodes and a single month can dominate
the whole series:

```{r}
idx  <- draws[[3]][[1]]
Sinv <- lapply(roll_S, function(S) solve(S[idx, idx, drop = FALSE]))
FC   <- roll_mu[, idx, drop = FALSE]
budget <- vapply(seq_along(oos), function(i) sum(Sinv[[i]] %*% FC[i, ]), numeric(1))
signif(quantile(abs(budget), c(0, 0.01, 0.05, 0.50)), 3)
```

Scale-invariant weights make the $\lambda$ signal reappear immediately: holding
$\lVert w\rVert_2 = 1$ instead of $\mathbf{1}^\top w = 1$, realised Sharpe rises
monotonically with $\lambda$ across the whole grid. The metrics are doing their
job either way; it is the leverage that is unstable.

This vignette reproduces the paper's *mechanism* on the paper's data and
configuration, at a reduced number of draws. It is not a substitute for the
replication package, which covers the market-timing and cross-sectional sorting
settings as well.

## Reference

Salcher, T., Stöckl, S., & Hanke, M. (2026). Lost in Translation?
Risk-Adjusting RMSE for Economic Forecast Performance. *Journal of Forecasting*.
\doi{10.1002/for.70134}
