---
title: "1. Evaluating forecasts: reproducing the published results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Evaluating forecasts: reproducing the published results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
options(digits = 4)
```

```{r}
library(rafe)
data(ff12)
```

This vignette reproduces the empirical results of Salcher, Stöckl & Hanke
(2026, *Journal of Forecasting*), Section 6, using only functions from this
package. At the end we check the numbers against the published table.

## The claim being tested

A forecast is scored with root mean squared error. A portfolio is scored with
its Sharpe ratio. If the first is a good proxy for the second, then forecasts
with lower RMSE should produce portfolios with a smaller shortfall against the
best attainable Sharpe ratio.

The paper's argument is that RMSE is what you get after stripping almost all
the risk information out of a sound error measure. Starting from RAFE and
imposing progressively cruder assumptions on the covariance matrix produces a
nested sequence ending in RMSE:

| Restriction | Covariance used to weight the error |
|---|---|
| `"none"` | the realised covariance $\Sigma$ — full RAFE |
| `"cc05"` | variances kept, all correlations set to $\rho = 0.5$ |
| `"cc0"` | variances kept, all correlations set to $0$ |
| `"cv"` | correlations $0$, all variances set to the average variance |
| `"i"` | the identity — this *is* RMSE |

`restrict_cov()` performs the restriction and every metric takes a `variant`
argument:

```{r}
set.seed(1)
S <- stats::cov(as.matrix(ff12[1:120, 2:6]))
round(stats::cov2cor(restrict_cov(S, "cc05")), 2)
restrict_cov(S, "i")
```

## The design

Monthly excess returns on the 12 Fama-French industry portfolios, January 1964
to December 2023. Rolling windows of 60 months for estimation and the following
36 months for evaluation, stepping one month at a time. The training moments
are the forecast; the realised moments of the test window are the truth.

```{r}
R     <- as.matrix(ff12[, -1])
train_len <- 60L
test_len  <- 36L
c(months = nrow(R), assets = ncol(R),
  windows = nrow(R) - train_len - test_len + 1L)
```

Portfolio weights use a ridge-shrunk covariance inverse, and the tangency
portfolio is normalised to full investment:

```{r}
ridge_inverse <- function(Rw, alpha = 0.20) {
  S  <- stats::cov(Rw)
  Ss <- (1 - alpha) * S + alpha * diag(diag(S))
  ev <- eigen(Ss, symmetric = TRUE, only.values = TRUE)$values
  if (min(ev) < 1e-6) Ss <- Ss + diag(abs(min(ev)) + 1e-4, ncol(Ss))
  solve(Ss)
}
tangency_weights <- function(mu, Sigma_inv) {
  w <- drop(Sigma_inv %*% mu); w / sum(w)
}
gmv_weights <- function(Sigma_inv) {
  w <- drop(Sigma_inv %*% rep(1, nrow(Sigma_inv))); w / sum(w)
}
realized_sharpe <- function(w, R_test) {
  r <- drop(R_test %*% w); mean(r) / stats::sd(r)
}
```

## The rolling evaluation

For each window we compute all five variants of RAFE, and the Sharpe-ratio gap
of two strategies: the tangency portfolio, whose perfect-foresight benchmark is
$SR^{*} = \sqrt{\mu^\top\Sigma^{-1}\mu}$, and the global minimum-variance
portfolio, whose benchmark uses the realised moments.

```{r}
variants <- c("none", "cc05", "cc0", "cv", "i")
starts   <- seq_len(nrow(R) - train_len - test_len + 1L)

panel <- do.call(rbind, lapply(starts, function(s) {
  tr <- R[s:(s + train_len - 1L), ]
  te <- R[(s + train_len):(s + train_len + test_len - 1L), ]

  mu_hat <- colMeans(tr)
  Sigma_w_inv <- ridge_inverse(tr)          # for weights
  mu    <- colMeans(te)
  Sigma <- stats::cov(te)
  Sigma_inv <- solve(Sigma)

  sr_real <- c(
    tan = realized_sharpe(tangency_weights(mu_hat, Sigma_w_inv), te),
    gmv = realized_sharpe(gmv_weights(Sigma_w_inv), te)
  )
  sr_oracle <- c(
    tan = sqrt(drop(t(mu) %*% Sigma_inv %*% mu)),
    gmv = realized_sharpe(gmv_weights(Sigma_inv), te)
  )

  errs <- vapply(variants, function(v) compute_rafe(mu_hat, mu, Sigma, variant = v),
                 numeric(1))
  c(errs, gap_tan = unname(sr_oracle["tan"] - sr_real["tan"]),
          gap_gmv = unname(sr_oracle["gmv"] - sr_real["gmv"]))
}))
panel <- as.data.frame(panel)
nrow(panel)
```

## Does the link to economic value survive the simplifications?

```{r}
tbl <- data.frame(
  metric = c("(none)", "-CC0.5", "-CC0", "-CV", "-I (RMSE)"),
  tan = vapply(variants, function(v) cor(panel[[v]], panel$gap_tan), numeric(1)),
  gmv = vapply(variants, function(v) cor(panel[[v]], panel$gap_gmv), numeric(1)),
  row.names = NULL
)
tbl$tan <- round(tbl$tan, 3)
tbl$gmv <- round(tbl$gmv, 3)
tbl
```

For the tangency portfolio the correlation between forecast error and economic
loss falls from **0.68** under the full RAFE to **0.11** once the covariance has
been stripped down to the identity. The same forecasts, the same portfolios, the
same Sharpe-ratio gaps — only the yardstick changes, and with it almost all of
the explanatory power.

```{r}
op <- par(mar = c(4.5, 4.5, 2.4, 1))
barplot(tbl$tan, names.arg = tbl$metric, col = "#1f4e79", border = NA,
        ylab = "correlation with the Sharpe-ratio gap",
        main = "Tangency portfolio: each simplification costs explanatory power")
abline(h = 0, col = "grey40")
par(op)
```

## Checking against the published table

The values above are Panel A of Table 4 in the paper. Reproducing them is the
sharpest available test that this package implements what was published:

```{r}
published <- data.frame(
  metric = c("(none)", "-CC0.5", "-CC0", "-CV", "-I (RMSE)"),
  tan = c(0.680, 0.440, 0.149, 0.190, 0.110),
  gmv = c(0.065, 0.126, 0.049, 0.041, 0.020)
)

comparison <- data.frame(
  metric        = published$metric,
  tan_package   = tbl$tan,
  tan_published = published$tan,
  gmv_package   = tbl$gmv,
  gmv_published = published$gmv
)
comparison

c(max_abs_difference = max(abs(c(tbl$tan - published$tan,
                                 tbl$gmv - published$gmv))))
```

Every entry agrees to the three decimal places the paper reports.

The full published analysis covers 13 portfolio strategies and a Monte Carlo
study of 100,000 paths alongside the empirical results; the replication code is
at
[github.com/sstoeckl/Lost_in_Translation_Replication](https://github.com/sstoeckl/Lost_in_Translation_Replication).
This vignette reproduces two of those strategies to keep the build fast.

## Reference

Salcher, L., Stöckl, S., & Hanke, M. (2026). Lost in Translation?
Risk-Adjusting RMSE for Economic Forecast Performance. *Journal of Forecasting*.
\doi{10.1002/for.70134}
