---
title: "0. rafe — getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{0. rafe — getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r}
library(rafe)
```

`rafe` implements decision-aligned forecast evaluation: metrics that score a
return forecast by the damage its errors do to a portfolio, rather than by
their size.

## The idea in one paragraph

Feed a forecast $(\hat\mu, \hat\Sigma)$ into a mean-variance optimiser and you
get a portfolio whose Sharpe ratio falls short of the oracle's by some gap
$\Delta$. Salcher, Stöckl & Hanke (2026) show that gap is bounded by two terms:
a mean error measured in the risk metric, and a covariance distortion measured
in the operator norm.

$$\Delta \;\le\; c \cdot \underbrace{\lVert\hat\mu - \mu\rVert_{\Sigma^{-1}}}_{\mathrm{RAFE}}
 \;+\; SR^{*} \cdot \underbrace{\lVert\Sigma^{1/2}\hat\Sigma^{-1}\Sigma^{1/2} - I\rVert_2}_{\mathrm{C\text{-}RAFE}}$$

RMSE appears nowhere. That is the point: it weights every asset equally, while
the optimiser does not.

## The metrics

```{r}
data(ff12)
R <- as.matrix(ff12[, -1])

train <- R[1:60, ]
eval  <- R[61:120, ]

mu_hat <- colMeans(train); Sigma_hat <- stats::cov(train)
mu     <- colMeans(eval);  Sigma     <- stats::cov(eval)

compute_rafe(mu_hat, mu, Sigma = Sigma)
compute_crafe(Sigma, Sigma_hat)
```

`compute_trafe()` combines them, applying the theorem's data-dependent constant
($c = 1$ if $\mathrm{RAFE} \le SR^{*}$, else $c = 2$) and returning the
decomposition as attributes:

```{r}
bound <- compute_trafe(mu_hat, mu, Sigma, Sigma_hat)
bound
unlist(attributes(bound))
```

## The correctors

If you cannot retrain the forecaster, correct its output. `sep_tune()` picks a
mean-shrinkage intensity $\kappa$ and a covariance eigenvalue floor $\tau$ on an
inner-validation split, then applies both:

```{r}
fit <- sep_tune(train, kappa_grid = seq(0, 1, by = 0.1),
                       tau_grid   = seq(0, 0.5, by = 0.1))
c(kappa = fit$kappa, tau = fit$tau_rel)

c(rafe_before  = compute_rafe(mu_hat, mu, Sigma = Sigma),
  rafe_after   = compute_rafe(fit$mu_tilde, mu, Sigma = Sigma),
  crafe_before = compute_crafe(Sigma, Sigma_hat),
  crafe_after  = compute_crafe(Sigma, fit$Sigma_tilde))
```

## Where to go next

- `vignette("rafe-evaluation", package = "rafe")` — the Part 1 evaluation
  example on the Fama-French 12 industries: does the bound hold, and does RAFE
  track economic loss where RMSE does not?
- `vignette("rafe-post-processing", package = "rafe")` — the Paper A
  post-processing example: tuning $(\kappa, \tau)$, what the correction does to
  both channels, and what it does to portfolios.

Training-loss objectives for XGBoost and torch (Paper B's lane) are not part of
this release.

## References

Salcher, T., Stöckl, S., & Hanke, M. (2026). Lost in Translation?
Risk-Adjusting RMSE for Economic Forecast Performance. *Journal of Forecasting*.
\doi{10.1002/for.70134}
